# Lets try harmony on the spatial data

Spatial data is very batch-effect-y. We need to correct for this. We can use the `harmony` package to do this.

I'm not convinced we need to run harmony here. Both samples were processes on the same visium chip and from the same experiment, but lets try it out. When I look at the UMAP, there are two clusters that correspond to the samples, but the spots with tumor all cluster together.

```{r get_data}
merged_data <- qs::qread("output/rdata/merged_data_w_col.qs")
```

### Lets plot the clusters before Harmony for later comparison
```{r plot_og_clusters}
og_plot_name <-
    SpatialDimPlot(merged_data,
                   ncol = 2,
                   pt.size.factor = 1,
                   label = TRUE)

og_dimplot <-
    DimPlot(merged_data,
            group.by = c("sample_id", "seurat_clusters"))
```

### Lets run Harmony and harmonize by the sample_id

We then re-process things using the harmony embeddings.

** Keep in mind that Harmony won't fix our DE analysis **
```{r run_harmony}
harm_data <-
    harmony::RunHarmony(merged_data, group.by.vars = "sample_id") %>%
    RunUMAP(reduction = "harmony", dims = 1:30) %>%
    FindNeighbors(dims = 1:30, reduction = "harmony") %>%
    FindClusters(resolution = 0.6)
```

### Lets make some plots to see how the harmony worked

```{r plot_harmony}
new_clusters <-
    SpatialDimPlot(harm_data,
                   ncol = 2,
                   pt.size.factor = 1,
                   label = TRUE)

combined_plot <- og_plot_name / new_clusters
ggsave("output/figures/harmony_spatial_plot.png",
       combined_plot,
       width = 15,
       height = 10)

new_dimplot <-
    DimPlot(harm_data,
            group.by = c("sample_id", "seurat_clusters"))

combined_dimplot <- og_dimplot / new_dimplot
ggsave("output/figures/harmony_spatial_dimplot.png",
       combined_dimplot,
       width = 15,
       height = 10)

plot_name <-
    DimPlot(harm_data,
            group.by = "high_col",
            split.by = "sample_id")
ggsave("output/figures/harmony_col1a1.png",
       plot_name,
       width = 15,
       height = 10)
```

It worked ok, I'm still not convinced. I think I'd have to go through and look at the cell type proportions to see if I believe that we need Harmony correction.