---
title: "Data Analysis of Spatial Test"
author: "Matt Cannon"
date: "`r format(Sys.time(), '%m/%d/%Y')`"
format:
    html:
        toc: true
        toc_float: true
        toc_depth: 5
        number_sections: false
        code-fold: true
        code-link: true
        df-print: kable
        self-contained: true
execute:
    cache: true
knitr:
    opts_chunk:
        cache.lazy: false
lightbox: true
---

# Sample summary

- 10x Visium run with cytassist
- 2 samples
    - Section with tumors
    - Section with less tumors

# Load packages

```{r lib}
library(tidyverse)
library(Seurat)
library(rrrSingleCellUtils)
theme_set(theme_bw())

# Set random generator seed to facilitate reproducibility
set.seed(888)
```

## Make up directory structure
```{bash mkdirs, eval=TRUE}
for directoryName in \
  output \
  output/figures \
  output/rdata \
  output/de
do
  if [ ! -d ${directoryName} ]
  then
    mkdir -p ${directoryName}
  fi
done
```

## Run spaceranger to generate fastq files

Make sure you correct your sample sheet before running spaceranger
```{bash spaceranger}
#| eval: false
sbatch spaceranger_mkfastq2.sh
```

## Run cellranger count to generate count matrices

```{bash cellranger}
#| eval: false
sbatch spaceranger_count.sh
```

## Read in all data and process it for downstream analysis
List of seurat object saved to `output/rdata/sobj_list.qs`

{{< include child_qmds/load_data.qmd >}}

## Lets try doing differential expression on the spots with lots of tumor

{{< include child_qmds/de_tumor.qmd >}}

## Look at how much the RNA difusses around outside of the tissue

{{< include child_qmds/test_diffusion.qmd >}}

## Run spot deconvolution to get cell type proportions for each spot

{{< include child_qmds/spacexr.qmd >}}

## Run harmony to remove batch effects

{{< include child_qmds/harmony.qmd >}}

```{r sessionInfo}
sessionInfo()
```
